\section{Introduction}\label{sec:intro}

Volatility forecasting is one of the few areas in finance where the question ``does this model actually work?'' has a relatively clean answer. Unlike return prediction, where the signal-to-noise ratio is so low that statistical significance is routinely debated, volatility is persistent, clustered, and measurable. A model that forecasts tomorrow's volatility well is immediately useful for option pricing, risk management, portfolio construction, and margin setting. The practical stakes are high, and the evaluation criteria are well-established.

What is less settled is which model works best. The field has accumulated a large zoo of approaches over the past three decades. On one end sit parametric time-series models: the GARCH family introduced by \citet{Bollerslev1986} and extended in dozens of directions (EGARCH, GJR-GARCH, FIGARCH, Realized GARCH, among others). In the middle is the Heterogeneous Autoregressive model for realized volatility (HAR-RV) of \citet{Corsi2009}, which achieves surprisingly good forecasting performance with nothing more than three lagged moving averages of realized volatility. On the other end sit machine learning methods: gradient-boosted trees, recurrent neural networks, and more recently, transformers and attention-based architectures.

Each new paper in this space typically compares its proposed method against one or two baselines and reports a win. The result is a literature that is hard to synthesize. Does LightGBM beat GARCH? It depends on the market, the sample period, the horizon, the loss function, and how carefully the author tuned the GARCH specification. Cross-study comparisons are close to meaningless because no two papers use the same data, features, or evaluation protocol.

This paper runs the horse race that the literature needs. We implement six models, comprising three GARCH variants (GARCH, EGARCH, GJR-GARCH), HAR-RV, LightGBM, and XGBoost, plus a simple equal-weight ensemble, and evaluate all of them on the same data, the same features, the same train/test split, using the same loss functions.\footnote{We initially planned to include LSTM and GRU recurrent networks. The PyTorch backend in our development environment had compatibility issues with the Python 3.14 toolchain we used elsewhere in the pipeline, and rather than introduce a separate environment for one model class we chose to defer the neural-network comparison. We flag this as a limitation in \Cref{sec:robustness}; the question of whether properly implemented sequence models add value over the methods we test here remains open.} The data is the S\&P 500, which we chose not because US equities are the most interesting market but because they are the most studied, making our results directly comparable to prior work. Our sample runs from January 2004 through November 2025, giving us 5,446 usable trading days after feature construction. The test period starts in January 2022, so every model must forecast through the 2022 bear market, the 2023 AI-driven rally, and the 2024--2025 rate normalization, a deliberately challenging out-of-sample environment for models trained on pre-pandemic data.

\subsection*{Why we wrote this paper}

Three concerns motivate the study, each addressing a habit in the existing literature that we think is worth changing.

\textit{Single-number reporting can mislead.} Most ML-vs-classical comparisons in this subfield report a full-sample QLIKE (or RMSE, or $R^2$) and stop. A reader has no way to tell whether the reported winner survives different regimes, holds up under statistical testing, or merely averages over heterogeneous environments that pull in different directions. We will show that the full-sample ranking in our own data hides a model-class reversal between 2022 and 2023--2025, and that several ostensible "wins" in the full-sample table do not survive a Diebold-Mariano test. We argue that any forecasting comparison in this space should report at minimum a full-sample table, a subperiod or regime-conditional table, and pairwise statistical significance.

\textit{Regime robustness, not average accuracy, is what risk managers buy.} The applied audience for these models is not a researcher checking a leaderboard; it is a risk manager who has to set initial margin tomorrow and a model-validation team that has to defend the choice next quarter. A model that wins by a wide margin in calm markets but fails in stressed ones is worse, in their use case, than a model that is never best but never disastrous either. Our test period (2022--2025) is constructed precisely to discriminate models on this property, and our conclusion that a simple equal-weight ensemble is the regime-robust choice is the result that matters most for that audience.

\textit{Reproducibility in this subfield is poor and we want to push back on that.} Cross-study comparisons in volatility-ML are nearly impossible because no two papers share data, features, or evaluation protocol. We release the entire pipeline (raw Yahoo Finance pull through final metric tables) under permissive licenses, scoring 2 on the Reproducibility Disclosure Score rubric proposed in \citet{Khan2026}. Anyone can rerun the comparison with a different feature set, an additional model, or a longer test period, and we explicitly invite such extensions through the repository's issue tracker.

\subsection*{Concrete contributions}

\begin{enumerate}[leftmargin=*]
    \item A head-to-head comparison of seven model configurations (three GARCH variants, HAR-RV, LightGBM, XGBoost, and an equal-weight ensemble) on post-pandemic US equity volatility data, evaluated on the same data, features, train/test split, and protocol.

    \item Three loss-function reports per model: QLIKE \citep{Patton2011}, RMSE, and Mincer-Zarnowitz $R^2$, accompanied by Diebold-Mariano two-sided $p$-values for all key model pairs with HAC (Newey-West) standard errors. Many ML papers in this space report only RMSE; we show that QLIKE and the DM test materially change the conclusions one can defend.

    \item A subperiod analysis along two cuts: calendar year (2022 versus 2023--2025) and realized-volatility regime (top quartile versus the lower three). The subperiod tables document a model-class reversal that the full-sample ranking hides.

    \item A demonstration that the equal-weight ensemble (LightGBM + HAR-RV + GARCH) is statistically indistinguishable from the best individual model on the full sample (DM $p = 0.90$ vs.\ GJR-GARCH) yet is more robust across regimes. We argue this is the practically useful version of the \emph{ensembles win} result.

    \item Full reproducibility: code, raw inputs, the engineered feature panel, the per-day forecasts of all seven model configurations, and the metric files used to populate every table in this paper are released at \href{https://github.com/ayk5511/volatility-forecasting}{\nolinkurl{github.com/ayk5511/volatility-forecasting}} under permissive licenses.
\end{enumerate}

The rest of the paper is organized as follows. \Cref{sec:literature} reviews the volatility forecasting literature with a focus on the ML methods introduced since 2015. \Cref{sec:data} describes our data and the construction of realized volatility measures. \Cref{sec:methodology} details each forecasting model and our evaluation protocol. \Cref{sec:results} presents the main results. \Cref{sec:robustness} reports the subperiod analysis and Diebold-Mariano significance tests, and discusses limitations. \Cref{sec:conclusion} concludes.
