\documentclass[12pt,a4paper]{article}

% === Packages ===
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{float}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=green!50!black,
    urlcolor=blue!70!black
}

\onehalfspacing

% === Title ===
\title{Volatility Forecasting with Machine Learning:\\A Horse Race Across GARCH, HAR, and Tree-Based Models}

\author{
    Akram Khan\thanks{Independent Researcher. ORCID: \href{https://orcid.org/0009-0002-7521-8648}{0009-0002-7521-8648}. Email: \href{mailto:1819ak@gmail.com}{1819ak@gmail.com}. SSRN: \href{https://ssrn.com/abstract=6663418}{ssrn.com/abstract=6663418}. Replication code, data, and results: \href{https://github.com/ayk5511/volatility-forecasting}{github.com/ayk5511/volatility-forecasting}.}
}

\date{April 2026}

\begin{document}

\maketitle

\begin{abstract}
This paper runs a controlled horse race across seven volatility-forecasting configurations on S\&P 500 realized volatility from January 2004 through November 2025. The seven are three GARCH variants (GARCH, EGARCH, GJR-GARCH), the HAR-RV model of \citet{Corsi2009}, two gradient-boosted trees (LightGBM, XGBoost), and an equal-weight ensemble of LightGBM, HAR-RV, and GARCH(1,1). Every model is evaluated on the same data, the same features, and the same expanding-window protocol over a January 2022--November 2025 test period of 980 trading days. The test window spans the 2022 bear market, the 2023 rally, the 2024--2025 rate-normalization regime, and a 19.4\% drawdown year. QLIKE serves as the primary loss function, alongside RMSE and Mincer-Zarnowitz $R^2$. Diebold-Mariano two-sided $p$-values are reported for all key pairs, plus the Hansen-Lunde-Nason Model Confidence Set, the Hansen Superior Predictive Ability test, and the White Reality Check, all computed with the open-source \texttt{vol-eval} package of \citet{Khan2026VolEval}.

Three findings emerge. First, on the full test sample no model separates from the field at conventional significance levels. The equal-weight ensemble has the lowest point-estimate QLIKE (0.3431); GJR-GARCH is the best individual model (QLIKE 0.3447); the Diebold-Mariano test cannot reject equal expected loss between them ($p = 0.90$). The ensemble's narrow lead is a precision result, not a power result. Second, the full-sample ranking masks a substantial subperiod reversal. In 2022 the GARCH family wins, with EGARCH best at QLIKE 0.2346 and LightGBM last. In 2023--2025 the tree models take the lead and HAR-RV drops to last. The 2022 environment is structurally out-of-distribution for models trained on 2004--2021, and the parametric GARCH variants adapt more gracefully than the 30-feature trees. Third, the LightGBM split-importance ranking shows VIX as the dominant feature; the leverage effect is captured through three different return-based features that together account for roughly 19\% of split count; and long-horizon RV signals dominate the short-horizon ones. The full LaTeX source, the data-collection and modeling scripts, the per-day forecasts of all seven models, and the supplementary metrics are released under permissive licenses, scoring 2 on the Reproducibility Disclosure Score rubric of \citet{Khan2026}.

\medskip
\noindent\textbf{Keywords:} volatility forecasting, realized volatility, GARCH, HAR-RV, machine learning, LightGBM, XGBoost, forecast combination, S\&P 500

\medskip
\noindent\textbf{JEL Classification:} C22, C45, C53, G17
\end{abstract}

\newpage
\tableofcontents
\newpage

% === Sections ===
\input{sections/01_introduction}
\input{sections/02_literature}
\input{sections/03_data}
\input{sections/04_methodology}
\input{sections/05_results}
\input{sections/06_robustness}
\input{sections/07_conclusion}

% === References ===
\bibliographystyle{apalike}
\bibliography{bib/references}

\end{document}
